---
title: "Exploratory Analytics and Time Series Modeliing on Petroleum product price Challenge"
author: "Abhinav Pathak"
output:
  pdf_document: default
  html_document: default
---


```{r}
suppressPackageStartupMessages({
  library(tidyr)
  library(dplyr)
  library(ggplot2)
  library(readxl)
  library(Quandl)
  library(gridExtra)
  library(TSA)
  library(forecast)
  library(sqldf)
})

```


### Loading data from excel file
```{r}
crude_oil_data <- read_excel("Jr._Data_Analyst_Project_File.xls", sheet = "Data 1", skip = 2) 
colnames(crude_oil_data) <- c("Date", "crude_cushing_WTI", "crude_brent_eu") 

conv_gasoln_data <- read_excel("Jr._Data_Analyst_Project_File.xls", sheet = "Data 2", skip = 2)
colnames(conv_gasoln_data) <- c("Date", "conv_gaso_ny", "conv_gaso_USgulf")

reg_gasoln_data <- read_excel("Jr._Data_Analyst_Project_File.xls", sheet = "Data 3", skip = 2)
colnames(reg_gasoln_data) <- c("Date", "reg_gas_la")

heat_oil_data <- read_excel("Jr._Data_Analyst_Project_File.xls", sheet = "Data 4", skip = 2)
colnames(heat_oil_data) <- c("Date", "heat_oil_ny")

sulfr_dsl_data <- read_excel("Jr._Data_Analyst_Project_File.xls", sheet = "Data 5", skip = 2)
colnames(sulfr_dsl_data) <- c("Date", "sulfr_dsl_ny","sulfr_dsl_USgulf","sulfr_dsl_la")

kersn_jet_data <- read_excel("Jr._Data_Analyst_Project_File.xls", sheet = "Data 6", skip = 2)
colnames(kersn_jet_data) <- c("Date", "kersn_jet_USgulf")

propn_data <- read_excel("Jr._Data_Analyst_Project_File.xls", sheet = "Data 7", skip = 2)
colnames(propn_data) <- c("Date", "propn_montBel")
```


### Putting all the data in one table

```{r}
crude_oil_data <- crude_oil_data %>% mutate(year = format(Date, "%Y"),
                          month = format(Date, '%m'))
heat_oil_data <- heat_oil_data %>% mutate(year = format(Date, "%Y"),
                          month = format(Date, '%m'))
propn_data <- propn_data %>% mutate(year = format(Date, "%Y"),
                          month = format(Date, '%m'))
reg_gasoln_data <- reg_gasoln_data %>% mutate(year = format(Date, "%Y"),
                          month = format(Date, '%m'))
conv_gasoln_data <- conv_gasoln_data %>% mutate(year = format(Date, "%Y"),
                          month = format(Date, '%m'))
sulfr_dsl_data <- sulfr_dsl_data %>% mutate(year = format(Date, "%Y"),
                          month = format(Date, '%m'))
kersn_jet_data <- kersn_jet_data %>% mutate(year = format(Date, "%Y"),
                          month = format(Date, '%m'))

petroleum_product_data <- crude_oil_data %>%
  full_join(heat_oil_data, by = c("month" = "month", "year" = "year")) %>%
  full_join(propn_data, by = c("month" = "month", "year" = "year")) %>%
  full_join(reg_gasoln_data, by = c("month" = "month", "year" = "year")) %>%
  full_join(conv_gasoln_data, by = c("month" = "month", "year" = "year")) %>%
  full_join(sulfr_dsl_data, by = c("month" = "month", "year" = "year")) %>%
  full_join(kersn_jet_data, by = c("month" = "month", "year" = "year")) %>%
  select(-Date.y, -Date.x.x, -Date.y.y, -Date.x.x.x, -Date.y.y.y,-Date)
  
```

# Question 1

### Visualizing the change in prices across time
```{r}
g1 <- ggplot(crude_oil_data) + 
  geom_line(aes(x= Date, y= crude_cushing_WTI, group =1 ), color = "red") + 
  geom_line(aes(x= Date, y= crude_brent_eu ), color = "blue") +
  ggtitle("crude oil prices across time") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

g2 <- ggplot(conv_gasoln_data) + 
  geom_line(aes(x= Date, y= conv_gaso_ny , group =1 ), color = "red") + 
  geom_line(aes(x= Date, y= conv_gaso_USgulf ), color = "blue") +
  ggtitle("Conventional gasoline prices across time")

g3 <- ggplot(reg_gasoln_data) + 
  geom_line(aes(x= Date, y= reg_gas_la , group =1 ), color = "red") +
  ggtitle("Regular gasoline prices across time")

g4 <- ggplot(heat_oil_data) + 
  geom_line(aes(x= Date, y= heat_oil_ny , group =1 ), color = "red") + 
  ggtitle("Heat oil prices across time")

g5 <- ggplot(sulfr_dsl_data) + 
  geom_line(aes(x= Date, y= sulfr_dsl_ny , group =1 ), color = "red") + 
    geom_line(aes(x= Date, y= sulfr_dsl_USgulf), color = "blue") + 
    geom_line(aes(x= Date, y= sulfr_dsl_la), color = "green") + 
  ggtitle("Sulfur Diesel prices across time")

g6 <- ggplot(kersn_jet_data) + 
  geom_line(aes(x= Date, y= kersn_jet_USgulf , group =1 ), color = "red") + 
  ggtitle("Kerosene Jet fuel prices across time")

g7 <- ggplot(propn_data) + 
  geom_line(aes(x= Date, y= propn_montBel , group =1 ), color = "red") +
  ggtitle("Propane fuel prices across time")

```

```{r}
g1
```

```{r}
g2
```

```{r}
g3
```

```{r}
g4
```

```{r}
g5
```

```{r}
g6
```

```{r}
g7
```


* Insight1: There is a similar trend acoss all types of oil prices. Since 2002, oil prices had been rising consistently.
* Insight2: Oil Prices reached its peak in July 2008 and prices were very high on average between 2008 and 2014
* Insight3: Prices have not reduced significantly compared to the average price of last 5-6 year

```{r}
month_wise_crude <- crude_oil_data %>% group_by(Date) %>% 
  mutate(year = format(Date,'%Y'),
         month = format(Date, '%m')) %>%
  group_by(month) %>%
  summarise(avg_crude_cushing_price = mean(crude_cushing_WTI))

ggplot(month_wise_crude) +
  geom_line(aes(x= factor(month), y = avg_crude_cushing_price, group = 1 )) + 
  labs(x = "Month", y = "Crude oil price", title = "Average crude oil price across months") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

* Insight4: Price of the crude oil is generally higher in summer season (June-July-August) and low in winters (Dec -Jan)

### Price across years 

```{r}
year_wise_crude <- crude_oil_data %>% group_by(Date) %>% 
  mutate(year = format(Date,'%Y'),
         month = format(Date, '%m')) %>%
  group_by(year) %>%
  summarise(avg_crude_cushing_price = mean(crude_cushing_WTI))

ggplot(year_wise_crude) +
    geom_line(aes(x= factor(year), y = avg_crude_cushing_price, group = 1 )) + 
    labs(x = "Year", y = "Crude oil price", title = "Average crude oil price across year") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

* Insight5: Price of petroleumn product was at its peak between 2008 and 2014


# Question 2

### Loading external data using Quandl package
```{r}

GDP_data <- Quandl("FRED/GDP", collapse = "monthly" ,start_date="1986-01-01", end_date="2017-01-01") %>% mutate(year = format(Date,'%Y'),month = format(Date,'%m'), GDP = Value) 
CPI_data = Quandl("FRED/CPIAUCSL",  collapse = "monthly" ,start_date="1986-01-01", end_date="2017-01-01") %>% mutate(year = format(Date,'%Y'),month = format(Date,'%m'), CPI = Value) 
Inflation_data = Quandl("WORLDBANK/USA_FP_CPI_TOTL_ZG",  collapse = "monthly" ,start_date="1986-01-01", end_date="2017-01-01") %>% mutate(year = format(Date,'%mz'),month = format(Date,'%m'), Inflation = Value) 


External_data <- GDP_data %>%
  full_join(CPI_data, by = c("month" = "month", "year"="year")) %>%
  full_join(Inflation_data, by = c("month" = "month", "year"="year"))

petroleum_product_data.external <- petroleum_product_data %>%
  full_join(External_data,  by = c("month" = "month", "year"="year"))

```

```{r}
ggplot(petroleum_product_data.external) +
  geom_point(aes(x =crude_cushing_WTI, y= GDP )) +
  ggtitle("crude oil vs GDP")
```

* Oil prices are correlated with the growth of economy

```{r}
ggplot(petroleum_product_data.external) +
  geom_point(aes(x =crude_cushing_WTI, y= CPI )) +
  ggtitle("crude oil vs Inflation")
```

* Oil prices are also increasin with consumer price  index. Therefore it is clear that external factors are also dependent in a way on oil prices although nothing can be said about causal relationships and it is just a correlation


# Question 3

## Price changes month over month

```{r}
crude_oil_data$crude_cushing_WTI_diff <- c(0, diff(crude_oil_data$crude_cushing_WTI, lag = 1))
ggplot(crude_oil_data) + 
  geom_line(aes(x= Date, y= crude_cushing_WTI_diff), color = "red") + 
  ggtitle("Price Change of crude oil prices across time")
```
```{r}

heat_oil_data$heat_oil_ny_diff <- c(0, diff(heat_oil_data$heat_oil_ny, lag = 1))
ggplot(heat_oil_data) + 
  geom_line(aes(x= Date, y= heat_oil_ny_diff), color = "red") + 
  ggtitle("Price Change of heat oil prices across time")

```

* Price has changed more abruptly from 2008 to 2015, whereas prior to 2008 the prices were a little stable

# Question 4

## Price changes and movement across year

```{r}
petroleum_product_data_byYear <- petroleum_product_data %>%
  group_by(year)  %>%
  summarise(crude_cushing_WTI = mean(crude_cushing_WTI, na.rm = TRUE),
            heat_oil_ny = mean(heat_oil_ny, na.rm = TRUE),
            propn_montBel =mean(propn_montBel, na.rm = TRUE),
            reg_gas_la = mean(reg_gas_la, na.rm = TRUE),
            conv_gaso_ny = mean(conv_gaso_ny, na.rm = TRUE),
            sulfr_dsl_ny = mean(sulfr_dsl_ny, na.rm = TRUE),
            kersn_jet_USgulf = mean(kersn_jet_USgulf, na.rm = TRUE))

h1 <- ggplot(petroleum_product_data_byYear) +
    geom_line(aes(x= factor(year), y = crude_cushing_WTI, group = 1 )) + 
    labs(x = "Year", y = "Crude oil price", title = "Average crude oil price across year") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

h2 <- ggplot(petroleum_product_data_byYear) +
    geom_line(aes(x= factor(year), y = heat_oil_ny, group = 1 )) + 
    labs(x = "Year", y = "heat oil ny", title = "Average heat oil price across year") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

h3 <- ggplot(petroleum_product_data_byYear) +
    geom_line(aes(x= factor(year), y = propn_montBel, group = 1 )) + 
    labs(x = "Year", y = "price", title = "Average propane price across year") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

h4 <- ggplot(petroleum_product_data_byYear) +
    geom_line(aes(x= factor(year), y = reg_gas_la, group = 1 )) + 
    labs(x = "Year", y = "price", title = "Average regular gasoline price across year") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

h5 <- ggplot(petroleum_product_data_byYear) +
    geom_line(aes(x= factor(year), y = conv_gaso_ny, group = 1 )) + 
    labs(x = "Year", y = "price", title = "Average conventional gasoline price across year") +   theme(axis.text.x = element_text(angle = 90, hjust = 1))

h6 <- ggplot(petroleum_product_data_byYear) +
    geom_line(aes(x= factor(year), y = sulfr_dsl_ny, group = 1 )) + 
    labs(x = "Year", y = "price", title = "Average sulfur diesel price across year") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

h7 <- ggplot(petroleum_product_data_byYear) +
    geom_line(aes(x= factor(year), y = kersn_jet_USgulf, group = 1 )) + 
    labs(x = "Year", y = "price", title = "Average kerosene jet price across year") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```



```{r}
h1
```

```{r}
h2
```

```{r}
h3
```

```{r}
h4
```

```{r}
h5
```

```{r}
h6
```

```{r}
h7
```
* Year wise trend has been similar for all the petroleum producrs with highers between 2008 and 2014

## Price movement

```{r}
petroleum_product_data_byYear_diff <- petroleum_product_data_byYear %>%
  mutate(crude_cushing_WTI = c(0, diff(crude_cushing_WTI, lag =1)),
            heat_oil_ny = c(0, diff(heat_oil_ny, lag =1)),
            propn_montBel = c(0, diff(propn_montBel, lag =1)),
            reg_gas_la = c(0, diff(reg_gas_la, lag =1)),
            conv_gaso_ny = c(0, diff(conv_gaso_ny, lag =1)),
            sulfr_dsl_ny = c(0, diff(sulfr_dsl_ny, lag =1)),
            kersn_jet_USgulf = c(0, diff(kersn_jet_USgulf, lag =1)))

d1 <- ggplot(petroleum_product_data_byYear_diff) +
    geom_line(aes(x= factor(year), y = crude_cushing_WTI, group = 1 )) + 
    labs(x = "Year", y = "price", title = "crude oil price movement by year") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

d2 <- ggplot(petroleum_product_data_byYear_diff) +
    geom_line(aes(x= factor(year), y = heat_oil_ny, group = 1 )) + 
    labs(x = "Year", y = "price", title = "heat oil price movement by year") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

d3 <- ggplot(petroleum_product_data_byYear_diff) +
    geom_line(aes(x= factor(year), y = propn_montBel, group = 1 )) + 
    labs(x = "Year", y = "price", title = "propane price movement by year") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

d4 <- ggplot(petroleum_product_data_byYear_diff) +
    geom_line(aes(x= factor(year), y = reg_gas_la, group = 1 )) + 
    labs(x = "Year", y = "price", title = "regular gasoline price movement across year") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

d5 <- ggplot(petroleum_product_data_byYear_diff) +
    geom_line(aes(x= factor(year), y = conv_gaso_ny, group = 1 )) + 
    labs(x = "Year", y = "price", title = "conventional gasoline price movement across year") +   theme(axis.text.x = element_text(angle = 90, hjust = 1))

d6 <- ggplot(petroleum_product_data_byYear_diff) +
    geom_line(aes(x= factor(year), y = sulfr_dsl_ny, group = 1 )) + 
    labs(x = "Year", y = "price", title = "sulfur diesel price movement across year") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

d7 <- ggplot(petroleum_product_data_byYear_diff) +
    geom_line(aes(x= factor(year), y = kersn_jet_USgulf, group = 1 )) + 
    labs(x = "Year", y = "price", title = "kerosene jet price movement across year") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
d1
```

```{r}
d2
```

```{r}
d3
```

```{r}
d4
```

```{r}
d5
```

```{r}
d6
```

```{r}
d7
```

* Price movement have been volatile in recent years fom 2008 to 2014

# Bonus questions

# Question 5

## Predict 6 months of Crude oil prices

```{r}
crude_oil <- petroleum_product_data %>% 
  select(crude_cushing_WTI) %>%
  filter(!is.na(crude_cushing_WTI))

ggAcf(crude_oil$crude_cushing_WTI)

```

```{r}
ggPacf(crude_oil$crude_cushing_WTI)
```

## check for stationarity
```{r}
adf.test(crude_oil$crude_cushing_WTI, alternative = "stationary")
```

* The series is stationary 
* From ACF and PACF plot, it is clear that this is AR2 process

```{r}
crude_oil.model <- Arima(crude_oil$crude_cushing_WTI,order=c(2,0,0))
      
```

```{r}
crude_oil.forecast <- forecast(crude_oil.model, h = 6)
crude_oil.forecast
```

```{r}
autoplot(crude_oil.forecast)
```

* According to the forecast, for the next 6 months crude oil prices will vary between 45 and 48


# Question 6

```{r}
ggplot(petroleum_product_data) + 
  geom_line(aes(x = Date.x, y = conv_gaso_ny), color = "blue",show.legend = ) + 
    geom_line(aes(x = Date.x, y = conv_gaso_USgulf), color = "red")
```

## find correlation
```{r}
ny <- petroleum_product_data$conv_gaso_ny
ny[is.na(ny)] <- 0
gulf <- petroleum_product_data$conv_gaso_USgulf
gulf[is.na(gulf)] <- 0
cor(ny,gulf)
```

* Correlation is very high between conventional gasoline prices between Ny and us gulf

# Question 7

```{r}
petroleum_product_data_byMonth <- petroleum_product_data %>% 
  group_by(month) %>%
  summarise(crude_cushing_WTI = mean(crude_cushing_WTI, na.rm = TRUE),
            heat_oil_ny = mean(heat_oil_ny, na.rm = TRUE),
            propn_montBel =mean(propn_montBel, na.rm = TRUE),
            reg_gas_la = mean(reg_gas_la, na.rm = TRUE),
            conv_gaso_ny = mean(conv_gaso_ny, na.rm = TRUE),
            sulfr_dsl_ny = mean(sulfr_dsl_ny, na.rm = TRUE),
            kersn_jet_USgulf = mean(kersn_jet_USgulf, na.rm = TRUE))

m1 <- ggplot(petroleum_product_data_byMonth) +
  geom_line(aes(x= month, y= crude_cushing_WTI, group = 1))
m2 <- ggplot(petroleum_product_data_byMonth) +
  geom_line(aes(x= month, y= heat_oil_ny, group = 1))
m3 <- ggplot(petroleum_product_data_byMonth) +
  geom_line(aes(x= month, y= propn_montBel, group = 1))
m4 <- ggplot(petroleum_product_data_byMonth) +
  geom_line(aes(x= month, y= reg_gas_la, group = 1))
m5 <- ggplot(petroleum_product_data_byMonth) +
  geom_line(aes(x= month, y= conv_gaso_ny, group = 1))
m6<- ggplot(petroleum_product_data_byMonth) +
  geom_line(aes(x= month, y= sulfr_dsl_ny, group = 1))
m7 <- ggplot(petroleum_product_data_byMonth) +
  geom_line(aes(x= month, y= kersn_jet_USgulf, group = 1))
```

```{r}
m1
m2
m3
m4
m5
m6
m7
```


* From all the 7 plots above for different petroleum products, it is clear that prices in December and January are generally lower than other months, Except in the case of propane prices, where is is lowest in the month of June


```{r}
petroleum_product_data_byQtr <- petroleum_product_data %>% 
  mutate (quarter = quarters(as.Date(Date.x))) %>%
  group_by(quarter) %>%
  summarise(crude_cushing_WTI = mean(crude_cushing_WTI, na.rm = TRUE),
            heat_oil_ny = mean(heat_oil_ny, na.rm = TRUE),
            propn_montBel =mean(propn_montBel, na.rm = TRUE),
            reg_gas_la = mean(reg_gas_la, na.rm = TRUE),
            conv_gaso_ny = mean(conv_gaso_ny, na.rm = TRUE),
            sulfr_dsl_ny = mean(sulfr_dsl_ny, na.rm = TRUE),
            kersn_jet_USgulf = mean(kersn_jet_USgulf, na.rm = TRUE))


q1 <- ggplot(petroleum_product_data_byQtr) +
  geom_line(aes(x= quarter, y= crude_cushing_WTI, group = 1))
q2 <- ggplot(petroleum_product_data_byQtr) +
  geom_line(aes(x= quarter, y= heat_oil_ny, group = 1))
q3 <- ggplot(petroleum_product_data_byQtr) +
  geom_line(aes(x= quarter, y= propn_montBel, group = 1))
q4 <- ggplot(petroleum_product_data_byQtr) +
  geom_line(aes(x= quarter, y= reg_gas_la, group = 1))
q5 <- ggplot(petroleum_product_data_byQtr) +
  geom_line(aes(x= quarter, y= conv_gaso_ny, group = 1))
q6<- ggplot(petroleum_product_data_byQtr) +
  geom_line(aes(x= quarter, y= sulfr_dsl_ny, group = 1))
q7 <- ggplot(petroleum_product_data_byQtr) +
  geom_line(aes(x= quarter, y= kersn_jet_USgulf, group = 1))

```

```{r}
q1
q2
q3
q4
q5
q6
q7
```

* Quarter 1 and Quarter 4 are generally the ones with the lowest prices, except propane prices where it is quarter 2 with lowest price

